---
output: html_document
---

```{r echo=FALSE,results='hide'}

```
# Exponential Distribution and Comparison with the Central Limit Theorem

Author: *Philip Abraham*  
Date: *October 5, 2016*


## Synopsis
The goal of the analysis was to determine whether the distribution of the means
and the variance of 1000 samples, each with 40 random exponentials, behave as 
predicted by the Central Limit Theorem. Based on the analysis done for the given 
exponential distribution, the sample statistics was found to be nearly normal, 
centered at the population mean, and the sample variance approximates to the 
square of the population standard deviation divided by the sample size. 
Therefore, this analysis validates the Central Limit Theorem.

## Central Limit Theorem
The Central Limit Theorem (CLT) states that the distribution of means of 
independent and identically distributed(iid) variables becomes that of a standard 
normal as the sample size increases.

## Analysis
In this project R program was utilized to see, empirically, how the Central Limit 
Theorem works. 

Forty iid exponentials, all with a rate parameter lambda = 0.2, were generated in R. 
This group of exponentials make up a sample of size 40. From this sample a statistic 
was computed. The statistic for this analysis is the mean of the sample (it varies 
for each sample). Samples each containing 40 iid exponentials, were generated 1000 
times in R, and the distribution of the 1000 sample means was investigated.

The goal of the analysis was to determine whether the distribution of the means,
and the variance of the 1000 samples, each with random 40 exponentials, behave 
as predicted by the Central Limit Theorem.

## Simulation
CLT states that, when certain conditions are satisfied, sample means are
approximately normally distributed, no matter what the underlying distribution is.

First, show a histogram (Figure 1) of the underlying distribution.
```{r echo=FALSE, fig.width=9, fig.height = 3, fig.align='center'}
# Figure 1 Code
set.seed(1234)
X <- rexp(1000,0.2)
hist(X, probability = TRUE,
     main = "Figure 1-1000 Exponential Distributions with a Rate(lambda) of 0.2",
     col="grey", xlab='Sample Means', breaks=100)
################################################################################
```
Looking at the distribution produced above in Figure 1, it's definitely not normal. 
It reflects an exponential distribution. The theoretical population mean($\mu$) 
and the population variance($\sigma^2$) for exponential distributions are given by:
$$\mu = \frac{1}{lambda} = \frac{1}{0.2} = 5$$ 

$$\sigma^2 = \frac{1}{lambda^2} = \frac{1}{0.2^2} = 25$$

However, the CLT is concerned with the distribution of means of the sample means.
Drawing 1000 samples (size = 40 exponential distribution each), a histogram of 
the distribution of sample means is generated, as shown below in Figure 2.
```{r echo=FALSE, fig.width=9, fig.height = 6, fig.align='center'}
# Code for generating distribution of sample means and the associated Figure 2
set.seed(112)
mns=NULL
for (i in 1 : 1000) mns = c(mns, mean(rexp(40,0.2)))
     hist(mns, probability = TRUE, 
main="Figure 2-Distribution of 1000 Sample Means of Exponentials-Sample Size(n) = 40 ", 
        xlab = "Means of Sample Means", breaks = 100, col ="grey")
abline(v=mean(mns), col="blue",lty=2, lwd=4)
abline(v=5, col="red",lty=2, lwd=3)
lines(density(mns), lwd=3, col="black")
text(3.8, 0.45, pos= 3.8, "Red line is the\n location of\n the theoretical mean at 5", 
     cex = .8, col = "red")
text(5.5, 0.58, pos= 4, 
"Blue line is the location\n of the calculated mean\n at 5.003\n of the sample means", 
     cex = .8, col = "blue")
################################################################################
```     

## Conclusions - Comparison of the results of the simulation with the CLT
For random samples of size n selected from some population with mean = $\mu$ and 
standard deviation = $\sigma$, the following three corollaries from the CLT hold:  

1. CLT states that the distribution of averages of iid variables becomes that of
a standard normal distribution as the sample size increases.

Looking at distribution produced above in Figure 2, while the original distribution 
was far from normal (see Figure 1), the distribution of sample means in Figure 2
is approximately normal.

A normal probability plot (NPP) was created below (Figure 3) with the sampling 
means distribution. NPP is a plot that determines if a sample is drawn 
from a normal distribution. If the data is drawn from a normal distribution, 
then the points will fall nearly in a straight line. The NPP shown below in 
Figure 3 indicates that sample means distribution follow a close approximation 
to a normal distribution with n = 40 sample size.
```{r echo=FALSE, fig.width=9, fig.height = 4.5, fig.align='center'}
# Code for the test of normality and the accompanying Figure 3 Q-Q Plot
     qqnorm(mns, main = "Figure 3-Normal Q-Q Plot, Sample Size n = 40");qqline(mns,col = 2)
################################################################################
```
2. The mean of the sample means and the population mean are equal.
$\bar X_n$ = $\mu$
```{r echo=FALSE}
# Code for calculating mean and variance of sampling mean distribution
X_n <- mean(mns) #Sample Mean
v <- var(mns) #Sample Variance
################################################################################
```
**The sample mean($\bar X_n$) generated in R for the sample mean distribution** 
**is `r round(X_n,3)`, which is very close to the calculated theoretical** 
**population mean($\mu$) of 5.** Figure 2 showing the distribution compares the 
locations of the theoretical mean and the sample means.

3. The population standard deviation($\sigma$) and the standard deviation of 
the sample means($\sigma_x$) are related by the formula:
$$\sigma_x = \frac{\sigma}{\sqrt{n}}$$
The theoretical sample variance(Var) is the square of the sample 
standard deviation($\sigma_x$).
$$Var = \sigma_x^2 =  \frac{\sigma^2}{n} = \frac{1}{lambda^2*n} = \frac{1}{(0.2^2)*(40)}$$
**Hence, the theoretical sample variance(Var) = 0.625.** Comparing with the
R calculated sample means distribution variance below:

Theoretical Var | Sample Means Dist.Var
-------------- | ----------------------
0.625         | `r round(v,3)`

 It shows that the variance (**`r round(v,3)`**) of this sample means distribution 
nearly approximates the theoretical population variance to within < 4%.

Based on the validation of the above three CLT corollaries, it can be concluded that 
the distribution of the means and the variance of the 1000 samples, each with 
random 40 exponentials, behave as predicted by the CLT.

## Appendix 
```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```

