---
output: pdf_document
---

```{r echo=FALSE,results='hide'}

```
# The Effect of the Delivery Method and/or Dosage Levels of Vitamin C on Tooth Growth in Guinea Pigs

Author: *Philip Abraham*  
Date: *October 8, 2016*

## Synopsis
Basic inferential data analysis was conducted with the ToothGrowth data
available in the R program data sets. The objective of the analysis was to
investigate whether the delivery method or the dosage levels of the supplement 
vitamin C, given to 60 independent guinea pig samples, either via orange juice or 
ascorbic acid affected the tooth growth lengths of guinea pigs.
This analysis concludes that even though the dosage levels of vitamin C DOES 
impact tooth growth lengths significantly, there is NOT a significant difference 
between tooth lengths for guinea pigs fed ascorbic and orange juice. 

## Data Set Information
The ToothGrowth data frame consists of 60 observations on 3 variables.
```{r echo=FALSE}
library(datasets)
data("ToothGrowth")
tooth_df <- ToothGrowth
str(tooth_df)
```

## Study Information
The data source C. I. Bliss (1952) "The Statistics of Biosassay"  cites the 
original study done by Crampton, E.W. **The Crampton paper indicates that these** 
**data are for 60 SEPARATE guinea pigs, hence was an INDEPENDENT samples study.**
Each of the 60 pigs received one of three dose levels of vitamin C (0.5, 1, and 
2 mg/day) by one of the two delivery methods - orange juice (OJ), or ascorbic 
acid (VC), a form of vitamin C. In essence, 30 guinea pigs were given orange juice,
and another 30 guinea pigs were given ascorbic acid.
The growth (len) of the odontoblasts of the incisor tooth was measured as a 
criterion of tooth growth in the guinea pigs.

## Exploratory Data Analysis
First, a plot was created to show the effect of dosage levels of orange juice(OJ) 
and ascorbic acid(VC) on tooth growth lengths in the guinea pigs. See Figure 1 below.
```{r echo=FALSE, fig.width=9, fig.height = 3, fig.align='center'}
# Code for Figure 1
library(ggplot2)
p <- ggplot(tooth_df,aes(dose,len, color=supp)) + 
scale_color_discrete(name="Legend") + 
theme(legend.key=element_rect(fill=NA)) + 
guides(colour = guide_legend(override.aes = list(size=5)))
#Add Box plot with jittered points
p1 <- p  + facet_grid(. ~ supp) + 
geom_jitter(shape=16, position=position_jitter(0.1),alpha=.5, size=5)
#Add trend line
p2 <- p1 + stat_summary(fun.y=mean, geom="line", size=3)
#Add titles and labels
p3<-p2 + 
labs(title="ToothGrowth data: length vs dose, given type of supplement", 
    x="Dosage Levels", y = "Tooth Growth Length")
p3+theme(panel.background = element_rect(fill = 'black'))
################################################################################
```
Figure 1 above shows a strong trend in the rise of tooth lengths with dosage levels
for guinea pigs fed either orange juice or ascorbic acid. 

Next a plot displaying pig tooth growth length by ascorbic acid(VC) and orange 
juice(OJ), given varying dosage levels is shown below in Figure 2.
```{r echo=FALSE, fig.width=9, fig.height = 3.5, fig.align='center'}
# Code for Figure 2
library(ggplot2)
p <- ggplot(tooth_df,aes(supp,len, color=supp)) + 
scale_color_discrete(name="Legend") + 
theme(legend.key=element_rect(fill=NA)) + 
guides(colour = guide_legend(override.aes = list(size=5)))
#Add Box plot with jittered points
p1 <- p + geom_boxplot(fill="darkolivegreen") + facet_grid(. ~ dose) + 
geom_jitter(shape=16, position=position_jitter(0.6),alpha=.8, size=2)
#Add means point for each delivery method
p2 <- p1 + stat_summary(fun.y=mean, geom="point", shape=23, size=8, 
color= "azure", fill="black")
#Add titles and labels
p3<-p2 + 
labs(title="Figure 2-ToothGrowth data: length vs supplement, given varying dosage levels", 
x="Delivery Method", y = "Tooth Growth Length")
# Change the panel color & yaxis limits
p3+theme(panel.background = element_rect(fill = 'black'))
################################################################################
```
Looking at Figure 2, it seems, at least visually, that delivering orange 
juice to the guinea pigs led to higher tooth growth in guinea pigs than delivering
ascorbic acid at the 0.5 and 1.0 mg/day levels. And on the 2.0 mg/day level
there doesn't seem to be a noticeable change in tooth growth for either 
delivery methods. So it might be interesting to investigate this relationship a 
bit further to to see if there is a statistically significant difference between 
tooth lengths for guinea pigs fed ascorbic versus guinea pigs fed orange juice. 

## Inferential Statistics
**Based on the trend shown in Figure 2 above, the null hypothesis($H_0$) for** 
**this analysis is the assertion that the delivery method of the supplement** 
**vitamin C, either via orange juice or ascorbic acid does effect tooth growth** 
**lengths in guinea pigs**. The alternate hypothesis($H_a$) will be that the 
there is no significant difference between tooth lengths for guinea pigs fed 
ascorbic and orange juice.
**A two-sample t-test was utilized to test the null hypothesis.** The t-test works 
well whenever the distribution of the data is roughly symmetric and mound shaped.
A histogram of the  data-set's numeric variable, len, shown below in Figure 3.
```{r echo=FALSE, fig.width=9, fig.height = 3, fig.align='center'}
# Code for Figure 3
hist(tooth_df$len, probability = TRUE,
     main = "Figure 3-Distribution of Tooth Lengths",
     col="darkorange", xlab='Tooth Lengths', breaks=6)
################################################################################
```
The histogram displays a slightly skewed distribution, **so a shapiro.test was**
**performed in R to test the degree of normality of the tooth growth length distribution.**
```{r echo=FALSE}
# Code for the shapiro.test
p<-shapiro.test(tooth_df$len)
p[2]
################################################################################
```
**The p-value (`r p[2]`) generated for the shapiro.test is >0.05, therefore it is** 
**assumed that the tooth growth length data has a Normal distribution.**

Next, a two-sample(independent groups) t-test was run in R, comparing the means 
of tooth growth lengths between the guinea pigs given orange juice and the guinea 
pigs on ascorbic acid.
```{r echo=FALSE}
# Code for the t.test for delivery method comparison
df_vc <- subset(tooth_df, tooth_df$supp == "VC")#guinea pigs on ascorbic acid
df_oj <- subset(tooth_df, tooth_df$supp == "OJ")#guinea pigs given orange juice
val1 <- t.test(df_oj$len, df_vc$len, paired = FALSE, var.equal = FALSE)
val1[3]
val1[4]
################################################################################
```
**With a p-value > 0.05 generated for the t-test, it can be concluded that there is** 
**no significant difference between tooth lengths for guinea pigs fed ascorbic acid** 
**and orange juice, The 95 percent confidence interval includes zero. Hence, the** 
**null hypothesis that these two groups have different tooth lengths is rejected.**

Figure 1 showed a strong trend in the rise of tooth lengths with dosage levels
for guinea pigs fed either orange juice or ascorbic acid.The relationship between 
tooth growth lengths and dosage levels (0.5 and 2.0 mg/day) of vitamin C was 
tested  in R with a two-sample(independent groups) t-test to see the statistical
significance of the relationship independent of the delivery method.
```{r echo=FALSE}
# Code for the t.test for dosage level comparison
df_5 <- subset(tooth_df, tooth_df$dose == 0.5)
df_2<- subset(tooth_df, tooth_df$dose == 2.0)
val2 <- t.test(df_2$len, df_5$len, paired = FALSE, var.equal = FALSE)
val2[3]
val2[4]
################################################################################
```
**A t-test run in R produced p-value << 0.05, and the 95 percent confidence** 
**interval does not include zero. This shows a strong correlation of tooth growth** 
**lengths with the dosage levels of vitamin C independent of whether the guinea** 
**pigs were given orange juice or ascorbic acid.**

## Appendix 
```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```

